A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-19 tex/codatadp.tex
more

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
%\newcommand{\Ec}{\mathsf{E}}
%\newcommand{\Ec}{E}
\newcommand{\Ec}{\mathcal{E}}
%\newcommand{\Fc}{\mathsf{F}}
%\newcommand{\Fc}{F}
\newcommand{\Fc}{\mathcal{F}}
%\newcommand{\Gc}{\mathsf{G}}
%\newcommand{\Ac}{\mathsf{A}}
%\newcommand{\Dc}{\mathsf{D}}
%\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\kern.45em \parallel \kern.45em}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{[#1]}
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\Var}{\mathcal{V}}
%\newcommand{\Varec}[1]{\Var{\ec{#1}}}
\newcommand{\Varec}[1]{\widetilde{#1}}
\newcommand{\VAREC}[1]{\widetilde{\vphantom{\scriptstyle x^i}\smash{#1}}}
\newcommand{\nf}[1]{{#1}\!\!\downarrow}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsim}{\aequiv}
\newcommand{\vsimv}[1]{\mathrel{=^{#1}_\alpha}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
\newcommand{\muvar}{\mathcal{Var}}

@@

% change?
\newcommand{\thO}{T_{\mathrm{o}}}
\newcommand{\thD}{T_{\mathrm{cd}}}

@@

\begin{theorem}[Termination]%
\label{thm:t}%
\afterDot
All derivation trees are finite.
\end{theorem}
\begin{proof}
Consider a derivation tree with root node $\Ec$.
Let $\SSS$ be the set of terms occurring as the argument of a selector in $\tEc$.
Let $T$ be the set of terms in $\tEc$ whose types are finite datatypes.
For each term $t \in T$,
let $\SSSS 0_t$ be the set $\{ t \}$,
for $i \geq 0$ let $\SSSS {i+1}_t =$ $\SSSS i_t \mathrel\cup \{ \const s( u ) \mid u : \upsilon \in \SSSS i, \upsilon \in \Data, \upsilon \text{ finite}, s \in \Sel^{\upsilon}  \}$,
and let $\SSSS \ast_t$ be the fixed point of this sequence.
We know this is a finite set for each $t$ since the values of the type of $t$ are of finite size.
Let $\SSSS \ast$ be the union of $\SSSS \ast_t$ for all $t \in T$,
and let $\tcEc$ be the set of subterms of $\Ec \mathrel\cup \{ \const C_i( \const s^1_i( t ), \ldots, \const s^{a_i}_i( t ) ) \mid t : \tau \in \SSS \mathrel\cup \SSSS \ast, \const C_i \in \Ctr^\tau \}$.
In a derivation tree with root node $\Ec$,
it can be shown by case analysis on the rules of the calculus that each non-root node $\Fc$ is such that
$\mathcal{T}(\Fc) \subseteq \tcEc$, and hence contains an equality between two terms from $\tcEc$ not occurring in its parent node.
Thus, the depth of a branch in a derivation tree with root node $\Ec$ is at most $\left| \tcEc \right|^2$,
which is finite since $\tcEc$ is finite.
\qed
\end{proof}

\begin{theorem}[Refutation Soundness]%
\label{thm:rs}%
\afterDot
If there exists a closed derivation tree with root node\/ $\Ec$, then\/ $\Ec$ is $\thD$-unsatisfiable.
\end{theorem}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Clash}, or \rn{Acyclic},
then $\Ec$ is $\thD$-unsatisfiable.
For \rn{Conflict}, this is a consequence of equality reasoning.
For \rn{Clash}, this is a consequence of distinctness.
For \rn{Acyclic}, our construction of $\Val$ indicates that the class of values that $t_1$ can take in models of $\Ec$ is infeasible,
and thus $\Ec$ is unsatisfiable.
If the children of $\Ec$ are closed derivation trees
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \mathrel\cup t \teq \const C_i( \const s^1_i( t ), \ldots, \const s^{a_i}_i( t ) )$ is unsatisfiable
for each $\const C_i \in \Ctr^\tau$.
Since by exhaustiveness, all models of $\thD$ entail (exactly) one equality $t \teq \const C_i( \const s^1_i( t ), \ldots, \const s^{a_i}_i( t ) )$,
we know $\Ec$ is $\thD$-unsatisfiable.
Otherwise, the child of $\Ec$ is a closed derivation tree
whose root node is $\Ec \mathrel\cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, \rn{Inject}, \rn{Unique}, or \rn{Singleton}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
For \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, this is a consequence of equality reasoning.
For \rn{Inject}, this is a consequence of injectivity.
For \rn{Unique}, our construction of $\Val$ indicates that the value of $t_1$ and $t_2$ are equivalent in all models of $\Ec$.
For \rn{Singleton}, clearly $t_1$ and $t_2$ must have the same value since the cardinality of their type is one.
By the induction hypothesis, we have $\Ec \mathrel\cup t_1 \teq t_2$ is $\thD$-unsatisfiable,
and thus $\Ec$ is $\thD$-unsatisfiable.
\qed
\end{proof}

@@

\begin{lemma}
\label{lem:mu-norm-arg}
If $\vec u$ are normal, then if $t = \nf{(\MU x.\; \const{C}( \vec u ) )} = \MU x.\; \const{C}( \vec w )$,
then $\vec u \vsim t\interp{\vec w}$.
\end{lemma}
\begin{proof}
%Since $u_i$ is normal, $w_i$ is the result of replacing all subterms $v$ of $u_i$ such that $t\interp{v} \aequiv t$ with $x$.
\rem{TODO}
\end{proof}

@@

To construct a completion of $\Val$, we must choose values for the free variables of $\Val$, ${\Varec{x}}$ and ${\Varec{w_3}}$.
The content of the set $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$ is $\alpha$-equivalent to
\[%\begin{array}{c}
\{
\MU x.\; \const{C}( \const{D}( x )),
\const{C}( \MU y.\; \const{D}( y )),
\MU x.\; \const{D}( x ),
\MU x.\; \const{E}( \const{C}( x )),
\MU x.\; \const{C}( \const{E}( x )),
\MU x.\; \const{C}( x )
\}
%\end{array}
\]
Consider a model $\M$ that interprets variables in $\Ec$ based on the mapping $\Val$,
for example, $x$ is interpretted as $\Val \ec{x}$, $y_1$ as $\Val \ec{y_1}$ and so on.
Assigning a value for ${\Varec{x}}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\M$ to be inconsistent.
For example, say we assign $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$ for ${\Varec{x}}$.
After this substitution, $\Val \ec{y_2}$ is $\MU {\Varec{y_2}}.\; \const{D}( \MU {\Varec{x}}.\; \const{D}( {\Varec{x}} ) )$,
which has normal form $\MU {\Varec{y_2}}.\; \const{D}( {\Varec{y_2}} )$,
which is $\alpha$-equivalent to $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$.
However, this contradicts the disequality $x \not\teq y_2$ in $\Ec$.
Inversely, if the value assigned to ${\Varec{x}}$ is not $\alpha$-equivalent to any term in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$,
then the values in the range of $\Val$ remain $\alpha$-disequivalent.
For example, we may assign values such as 
$\MU {\Varec{x}}.\; \const{E}( {\Varec{x}} )$,
$\MU {\Varec{x}}.\; \const{E}( \MU x_1.\; \const{C}( x_1) )$, or
$\MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) )$
to ${\Varec{x}}$.
Legal substitutions for $v_{\ec{x}}$ may cause the range of $\Val$ to contain non-normal terms.
After applying the substitution for the third term mentioned above,
$\Val \ec{w_1}$ is $\MU {\Varec{w_1}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) ) )$,
which has normal form
$\MU {\Varec{w_1}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( {\Varec{w_1}} ) )$.
\xend
\end{examplex}

@@

\begin{lemma}
\label{lem:model-completion}
If $\Val$ is constructed for a saturated $\Ec$,
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, then:
\begin{enumerate}
\item $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$, where $\ec{x}$ has type $\tau$,
\item
All subterms $\MU {\Varec{x}}.\; t$ of $\Val^\ast \ec{y}$ are such that
$\Val^\ast \ec{y} \interp{ \MU {\Varec{x}}.\; t } = \Val^\ast \ec{x}$. 
%If $\const{C}( t_1, \ldots, t_n ) \in \ec{x}$,
%then all subterms $\MU v_{\ec{x}}.\; u$ occurring in $\Val^\ast \ec{y}$ 
%are of the form $\MU v_{\ec{x}}.\; \const{C}( w_1, \ldots, w_n )$,
%where 
%and $\Val^\ast \ec{y} \interp{ w_i } = \Val^\ast \ec{t_i}$ for $i = 1, \ldots, n$.
%If $\Val^\ast \ec{x}$ contains a subterm of the form $\MU v_{\ec{x}}.\; \const{C}( w_1, \ldots, w_n )$,
%and $\muvar( w_i ) = v_{\ec{t_i}}$ for $i = 1, \ldots, n$,
%\item
%$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
%for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)={\Varec{x}}$,
\item
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_x(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to ${\Varec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction,
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to ${\Varec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$.

@@

If $t$ is a selector term of the form $\const s^j_k( u )$,
since \rn{Split} does not apply,
$\ec{u}$ must contain a term of the form $\const C_i( \const s^1_i( u ), \ldots, \const s^{n}_i( u ) )$ for some $i$.
By Lemma~\ref{lem:model-completion}.2 and due to our construction of $\Val^\ast$, we have $\Val^\ast \ec{u}$ is a term $t_w$ of the form
$\MU {\Varec{u}}.\; \const C_i( w_1, \ldots, w_n )$,
where $t_w \interp{ w_j } = \Val^\ast \ec{ \const s^j_k( u )}$. 
If $k = i$, then $\M( t )$ is $\alpha$-equivalent with $t_w\interp{w_j} = \Val^\ast \ec{ \const s^j_k( u )}$,
which is $\Val^\ast \ec{t}$.
If $k \neq i$, then we interpret $\const s^j_k( \M( u ) )$ as the value in $\M( \tau )$ that is $\alpha$-equivalent with $\Val^\ast \ec{ t }$.
Since \rn{Cong} does not apply and due to Lemma~\ref{lem:model-completion}.3,
pairs of wrongly-applied selectors are assigned the same value if and only if they reside in the same equivalence class,
and thus this definition is well-defined.

@@

Discriminators are treated as predicate symbols, so a discriminator $\const{d}_i( t )$ is such that $\const{d}_i$
is a predicate symbol, instead of reducing this constraint to $t \teq \const{C}_i( \const s^1_i( t ), \ldots, \const s^n_i( t ) )$.
We have found that this leads to better performance since this reduction introduces terms more eagerly to $\tEc$.
Handling discriminators requires extending the decision procedure with several additional rules (see~\cite{} for more details), which apply uniformly to both datatypes and codatatypes.
Also, selectors are collapsed eagerly:
if $\const s^j_i( t )$ is a term in $\tEc$ and $t$ contains the constructor
$\const{C}_i( \vec u )$, then we immediately infer $\const s^j_i( t ) \teq u_j$, whereas the calculus we presented would apply \rn{Split} and \rn{Inject} before inferring this equality.
For the sake of reducing the number of unique constraints considered by the calculus, we compute a normal form for constraints as a preprocessing step.
In particular, we
replace $t \teq s$ by $s \teq t$ if $s$ is less than $t$ based on a term ordering,
replace $\const{C}_i( \vec t ) \teq \const{C}_j( \vec u )$ with $\bot$ when $i \neq j$,
replace all selector terms of the form $\const{s}^{\,i}_j( \const{C}_j(t_1,\ldots,t_n) )$ by $t_i$,
and replace all occurrences of discriminators $\const{d}_i( \const{C}_k( \vec t ) )$ by $\top$ if $i=k$ and $\bot$ if $i \neq k$.

\paragraph{Integration into DPLL(T)}
The calculus is implemented as \emph{theory solver} of CVC4,
that is, a specialized solver for determining the satisfiability of conjunctions of constraints for its theory.
Given a theory $T = T_1 \mathrel\cup \ldots T_n$ and a input set of clauses $F$ in CNF,
the DPLL(T) \rem{CDCL(T)?} procedure
(incrementally) builds partial assignments from the atomic formulas of $F$ to truth values such that no clause in $F$ is falsified.
We may interpret this partial assignment as a set of literals where $a \in M$ for all atoms $a$ that it assigned the value $\top$, and $\neg a \in M$ for all atoms that it assigns the value $\bot$.
Then, at a high level, by the Nelson--Oppen combination framework,
each $T_i$-solver for $i = 1, \ldots, n$ takes as input a combination of (1) the purified form of $T_i$-constraints occurring in $M$ where terms containing symbols not belonging to $T_i$ are replaced by fresh variables of the same type,
(2) additional equalities and disequalities between variables of types not belonging to $T_i$.
Let us call this set $M_i$.
Each $T_i$ solver either
reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case $\neg C$ is added as a additional clause to $F$,
adds an additional clause $C$ to $F$,
or does nothing.
When $M$ is a complete assignment for $F$, then a theory solver can only choose to do nothing when $M_i$ is indeed $T_i$-satisfiable.

>>>>

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
%\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Ec}{E}
%\newcommand{\Ec}{\mathcal{E}}
%\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Fc}{F}
%\newcommand{\Fc}{\mathcal{F}}
%\newcommand{\Gc}{\mathsf{G}}
%\newcommand{\Ac}{\mathsf{A}}
%\newcommand{\Dc}{\mathsf{D}}
%\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\kern.45em \parallel \kern.45em}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{[#1]}
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\Var}{\mathcal{V}}
%\newcommand{\Varec}[1]{\Var{\ec{#1}}}
\newcommand{\Varec}[1]{\widetilde{#1}}
\newcommand{\VAREC}[1]{\widetilde{\vphantom{\scriptstyle x^i}\smash{#1}}}
\newcommand{\nf}[1]{{#1}\!\!\downarrow}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsim}{\aequiv}
\newcommand{\vsimv}[1]{\mathrel{=^{#1}_\alpha}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
\newcommand{\muvar}{\mathcal{Var}}

@@

% change?
\newcommand{\thO}{T_{\mathrm{o}}}
\newcommand{\thD}{T_{\!\mathrm{cd}}}

@@

\begin{theorem}[Termination]%
\label{thm:t}%
\afterDot
All derivation trees are finite.
\end{theorem}
\begin{proof}
Consider a derivation tree with root node $\Ec$.
Let $\SSS$ be the set of terms occurring as the argument of a selector in $\tEc$.
Let $T$ be the set of terms in $\tEc$ whose types are finite datatypes.
For each term $t \in T$,
let $\SSSS 0_t$ be the set $\{ t \}$,
for $i \geq 0$ let $\SSSS {i+1}_t =$ $\SSSS i_t \mathrel\cup \{ \const s( u ) \mid u : \upsilon \in \SSSS i, \upsilon \in \Data, \upsilon \text{ finite}, s \in \Sel^{\upsilon}  \}$,
and let $\SSSS \ast_t$ be the fixed point of this sequence.
We know this is a finite set for each $t$ since the values of the type of $t$ are of finite size.
Let $\SSSS \ast$ be the union of $\SSSS \ast_t$ for all $t \in T$,
and let $\tcEc$ be the set of subterms of $\Ec \mathrel\cup \{ \const C_i( \const s^1_i( t ), \ldots, \const s^{a_i}_i( t ) ) \mid t : \delta \in \SSS \mathrel\cup \SSSS \ast,\, \const C_i \in \Ctr^\delta \}$.
In a derivation tree with root node~$\Ec$,
it can be shown by case analysis on the rules of the calculus that each non-root node $\Fc$ is such that
$\mathcal{T}(\Fc) \subseteq \tcEc$, and hence contains an equality between two terms from $\tcEc$ not occurring in its parent node.
Thus, the depth of a branch in a derivation tree with root node $\Ec$ is at most $\left| \smash{\tcEc}\vphantom{X} \right|{\!}^2$,
which is finite since $\tcEc$ is finite.
\qed
\end{proof}

\begin{theorem}[Refutation Soundness]%
\label{thm:rs}%
\afterDot
If there exists a closed derivation tree with root node\/ $\Ec$, then\/ $\Ec$ is $\thD$-unsatisfiable.
\end{theorem}
\begin{proof}
The proof is by structural induction on the derivation tree with root node $\Ec$.
If the tree is an application of \rn{Conflict}, \rn{Clash}, or \rn{Acyclic},
then $\Ec$ is $\thD$-unsatisfiable.
For \rn{Conflict}, this is a consequence of equality reasoning.
For \rn{Clash}, this is a consequence of distinctness.
For \rn{Acyclic}, the construction of $\Val$ indicates that the class of values that $t_1$ can take in models of $\Ec$ is infeasible,
and thus $\Ec$ is unsatisfiable.
If the children of $\Ec$ are closed derivation trees
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\delta$,
by the induction hypothesis $\Ec \mathrel\cup t \teq \const C_i( \const s^1_i( t ), \ldots, \const s^{a_i}_i( t ) )$ is unsatisfiable
for each $\const C_i \in \Ctr^\delta$.
Since by exhaustiveness, all models of $\thD$ entail (exactly) one equality $t \teq \const C_i( \const s^1_i( t ), \ldots, \const s^{a_i}_i( t ) )$,
we know $\Ec$ is $\thD$-unsatisfiable.
Otherwise, the child of $\Ec$ is a closed derivation tree
whose root node is $\Ec \mathrel\cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, \rn{Inject}, \rn{Unique}, or \rn{Singleton}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
For \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, this is a consequence of equality reasoning.
For \rn{Inject}, this is a consequence of injectivity.
For \rn{Unique}, our construction of $\Val$ indicates that the value of $t_1$ and $t_2$ are equivalent in all models of $\Ec$.
For \rn{Singleton}, clearly $t_1$ and $t_2$ must have the same value since the cardinality of their type is one.
By the induction hypothesis, we have $\Ec \mathrel\cup t_1 \teq t_2$ is $\thD$-unsatisfiable,
and thus $\Ec$ is $\thD$-unsatisfiable.
\qed
\end{proof}

@@

\begin{lemma}
\label{lem:mu-norm-arg}
If $\vec u$ are normal and $t = \nf{(\MU x.\; \const{C}( \vec u ) )} = \MU x.\; \const{C}( \vec w )$,
then $\vec u \vsim t\interp{\vec w}$.
\end{lemma}
\begin{proof}
%Since $u_i$ is normal, $w_i$ is the result of replacing all subterms $v$ of $u_i$ such that $t\interp{v} \aequiv t$ with $x$.
\rem{TODO}
\end{proof}

@@

To construct a completion of $\Val$, we must choose values for the free variables of $\Val$, ${\Varec{x}}$ and ${\Varec{w_3}}$.
The content of the set $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$ is $\alpha$-equivalent to
\[%\begin{array}{c}
\{
\MU x.\; \const{C}( \const{D}( x )),
\const{C}( \MU y.\; \const{D}( y )),
\MU x.\; \const{D}( x ),
\MU x.\; \const{E}( \const{C}( x )),
\MU x.\; \const{C}( \const{E}( x )),
\MU x.\; \const{C}( x )
\}
%\end{array}
\]
Consider a model $\M$ that interprets variables in $\Ec$ based on the mapping $\Val$,
for example, $x$ is interpretted as $\Val \ec{x}$, $y_1$ as $\Val \ec{y_1}$ and so on.
Assigning a value for ${\Varec{x}}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\M$ to be inconsistent.
For example, say we assign $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$ for ${\Varec{x}}$.
After this substitution, $\Val \ec{y_2}$ is $\MU {\Varec{y_2}}.\; \const{D}( \MU {\Varec{x}}.\; \const{D}( {\Varec{x}} ) )$,
which has normal form $\MU {\Varec{y_2}}.\; \const{D}( {\Varec{y_2}} )$,
which is $\alpha$-equivalent to $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$.
However, this contradicts the disequality $x \not\teq y_2$ in $\Ec$.
Inversely, if the value assigned to ${\Varec{x}}$ is not $\alpha$-equivalent to any term in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$,
the values in the range of $\Val$ remain $\alpha$-disequivalent.
For example, we may assign values such as 
$\MU {\Varec{x}}.\; \const{E}( {\Varec{x}} )$,
$\MU {\Varec{x}}.\; \const{E}( \MU x_1.\; \const{C}( x_1) )$, or
$\MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) )$
to ${\Varec{x}}$.
Legal substitutions for $v_{\ec{x}}$ may cause the range of $\Val$ to contain non-normal terms.
After applying the substitution for the third term mentioned above,
$\Val \ec{w_1}$ is $\MU {\Varec{w_1}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) ) )$,
which has normal form
$\MU {\Varec{w_1}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( {\Varec{w_1}} ) )$.
\xend
\end{examplex}

@@

\begin{lemma}
\label{lem:model-completion}
If $\Val$ is constructed for a saturated $\Ec$,
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, then
\begin{enumerate}
\item $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$, where $\ec{x}$ has type $\tau$,
\item
All subterms $\MU {\Varec{x}}.\; t$ of $\Val^\ast \ec{y}$ are such that
$\Val^\ast \ec{y} \interp{ \MU {\Varec{x}}.\; t } = \Val^\ast \ec{x}$. 
%If $\const{C}( t_1, \ldots, t_n ) \in \ec{x}$,
%then all subterms $\MU v_{\ec{x}}.\; u$ occurring in $\Val^\ast \ec{y}$ 
%are of the form $\MU v_{\ec{x}}.\; \const{C}( w_1, \ldots, w_n )$,
%where 
%and $\Val^\ast \ec{y} \interp{ w_i } = \Val^\ast \ec{t_i}$ for $i = 1, \ldots, n$.
%If $\Val^\ast \ec{x}$ contains a subterm of the form $\MU v_{\ec{x}}.\; \const{C}( w_1, \ldots, w_n )$,
%and $\muvar( w_i ) = v_{\ec{t_i}}$ for $i = 1, \ldots, n$,
%\item
%$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
%for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)={\Varec{x}}$,
\item
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_x(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to ${\Varec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction,
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to ${\Varec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$.

@@

If $t$ is a selector term of the form $\const s^j_k( u )$,
since \rn{Split} does not apply,
$\ec{u}$ must contain a term of the form $\const C_i( \const s^1_i( u ), \ldots, \const s^{n}_i( u ) )$ for some $i$.
By Lemma~\ref{lem:model-completion}.2 and due to our construction of $\Val^\ast$, we have $\Val^\ast \ec{u}$ is a term $t_w$ of the form
$\MU {\Varec{u}}.\; \const C_i( w_1, \ldots, w_n )$,
where $t_w \interp{ w_j } = \Val^\ast \ec{ \const s^j_k( u )}$. 
If $k = i$, then $\M( t )$ is $\alpha$-equivalent with $t_w\interp{w_j} = \Val^\ast \ec{ \const s^j_k( u )}$,
which is $\Val^\ast \ec{t}$.
If $k \neq i$, we interpret $\const s^j_k( \M( u ) )$ as the value in $\M( \tau )$ that is $\alpha$-equivalent with $\Val^\ast \ec{ t }$.
Since \rn{Cong} does not apply and due to Lemma~\ref{lem:model-completion}.3,
pairs of wrongly-applied selectors are assigned the same value if and only if they reside in the same equivalence class,
and thus this definition is well-defined.

@@

Discriminators are treated as predicate symbols, so a discriminator $\const{d}_i( t )$ is such that $\const{d}_i$
is a predicate symbol, instead of reducing this constraint to $t \teq \const{C}_i( \const s^1_i( t ), \ldots, \const s^n_i( t ) )$.
We have found that this leads to better performance since this reduction introduces terms more eagerly to $\tEc$.
Handling discriminators requires extending the decision procedure with several additional rules (see~\cite{} for more details), which apply uniformly to both datatypes and codatatypes.
Also, selectors are collapsed eagerly:
if $\const s^j_i( t )$ is a term in $\tEc$ and $t$ contains the constructor
$\const{C}_i( \vec u )$, we immediately infer $\const s^j_i( t ) \teq u_j$, whereas the calculus we presented would apply \rn{Split} and \rn{Inject} before inferring this equality.
For the sake of reducing the number of unique constraints considered by the calculus, we compute a normal form for constraints as a preprocessing step.
In particular, we
replace $t \teq s$ by $s \teq t$ if $s$ is less than $t$ based on a term ordering,
replace $\const{C}_i( \vec t ) \teq \const{C}_j( \vec u )$ with $\bot$ when $i \neq j$,
replace all selector terms of the form $\const{s}^{\,i}_j( \const{C}_j(t_1,\ldots,t_n) )$ by $t_i$,
and replace all occurrences of discriminators $\const{d}_i( \const{C}_k( \vec t ) )$ by $\top$ if $i=k$ and $\bot$ if $i \neq k$.

\paragraph{Integration into DPLL(T)}
The calculus is implemented as \emph{theory solver} of CVC4,
that is, a specialized solver for determining the satisfiability of conjunctions of constraints for its theory.
Given a theory $T = T_1 \mathrel\cup \ldots T_n$ and a input set of clauses $F$ in CNF,
the DPLL(T) \rem{CDCL(T)?} procedure
(incrementally) builds partial assignments from the atomic formulas of $F$ to truth values such that no clause in $F$ is falsified.
We may interpret this partial assignment as a set of literals where $a \in M$ for all atoms $a$ that it assigned the value $\top$, and $\neg a \in M$ for all atoms that it assigns the value $\bot$.
Then, at a high level, by the Nelson--Oppen combination framework,
each $T_i$-solver for $i = 1, \ldots, n$ takes as input a combination of (1) the purified form of $T_i$-constraints occurring in $M$ where terms containing symbols not belonging to $T_i$ are replaced by fresh variables of the same type,
(2) additional equalities and disequalities between variables of types not belonging to $T_i$.
Let us call this set $M_i$.
Each $T_i$ solver either
reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case $\neg C$ is added as a additional clause to $F$,
adds an additional clause $C$ to $F$,
or does nothing.
When $M$ is a complete assignment for $F$, a theory solver can only choose to do nothing when $M_i$ is indeed $T_i$-satisfiable.
